Local Spectral Clustering of Density Upper Level Sets

### 2019-05-23 neurips/local_spectral_density_clustering_neurips.tex
Through Sect 2

When applied to geometric graphs built from a large number of samples,
global spectral clustering methods can be computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to recent increased interest in
local spectral algorithms, which leverage locally-biased spectra computed using
random walks around a user-specified seed node.  A popular local clustering
algorithm is Personalized PageRank (PPR), first introduced by
\citet{haveliwala2003}, and then further developed by
\citep{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013},
among others.

@@

\paragraph{Estimation of density clusters} Let \smash{$\Cbb_f(\lambda)$} denote
the connected components of the density upper level set $\{x \in \Rd: f(x) >
\lambda\}$.  For a given density cluster \smash{$\Cset \in \Cbb_f(\lambda)$}, we
call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical density cluster}. The
size of the symmetric set difference between estimated and empirical clusters is 
a commonly used metric to quantify cluster estimation error
\citep{korostelev1993,polonik1995,rigollet2009}.

@@

Note that this symmetric set difference metric does not account for the distance
points in \smash{$\Cest \setminus \mathcal{S}[\Xbf]$} may be from
$\mathcal{S}$  \citep{singh2009}. Hence in this paper, in addition to the symmetric
set difference metric, we also study a second asymptotic notion for the quality
of a cluster estimator introduced by \citet{hartigan1981}.

% which measures whether $\Cest$ can distinguish any two distinct
% elements $\Cset, \Cset' \in \Cbb_f(\lambda)$.

@@

More broadly, density clustering and level set estimation is a well-studied
problem. \citet{polonik1995, rigollet2009} study density clustering under
symmetric set difference, \citet{tsybakov1997, singh2009} exhibit minimax
optimal level set estimators under Hausdorff loss and \citet{hartigan1981,
  chaudhuri2010} consider consistent estimation of the cluster tree, to note but
a few works in the area. Our goal is not to improve on these results, nor to 
offer a new algorithm for level set estimation; indeed, seen as a density
clustering algorithm, PPR has none of the optimality guarantees of the
methods in the aforementioned works. Rather, our motivation is start with a 
popular  and flexible local spectral algorithm, PPR, and to better understand
and characterize the distinctions between those density clusters which are 
well-conditioned for PPR, and those which are not.

@@

We formalize some geometric conditions, before using these to define a condition number $\kappa(\Cset)$ which measures the difficulty \pprspace will have in estimating $\Cset$. We motivate this measure, and the underlying geometric conditions, by giving density cluster estimation guarantees for Algorithm \ref{alg: ppr} in terms of $\kappa(\Cset)$.

\paragraph{Geometric conditions on density clusters}

As mentioned previously, successful recovery of a density cluster by \pprspace requires the density cluster to be geometrically well-conditioned. At a minimum, we wish to avoid sets $\Cset$ which contain arbitrarily thin bridges or spikes, and therefore as in \cite{chaudhuri2010} we introduce a buffer zone around $\Cset$. Let $B(x,r)$ be the closed ball of radius $r > 0$ centered at $x \in \Rd$. For $\lambda > 0$, consider a given cluster $\Cset \in \Cbb_f(\lambda)$. We denote the distance between $x$ and $\Cset$ as $\dist(x,\Cset) := \inf_{y \in \Cset} \norm{y - x}$, and for a given $\sigma > 0$, we refer to $\Csig := \set{x \in \Reals^d: \dist(x,\Cset) \leq \sigma}$ as the $\sigma$-expansion of $\Cset$. We now state our conditions with respect to $\Csig$.

\begin{enumerate}[label=(A\arabic*)]
	\item
	\label{asmp: bounded_density}
	\emph{Bounded density within cluster:} There exist constants $\lambda_{\sigma}, \Lambda_{\sigma}$ such that $0 < \lambda_{\sigma} = \inf_{x \in \Csig} f(x) \leq \sup_{x \in \Csig} f(x) \leq \Lambda_{\sigma} < \infty$.

	\item
	\label{asmp: cluster_separation}
	\emph{Cluster separation:}
	For all $\Cset' \in \Cbb_f(\lambda)$ with $\Cset' \not= \Cset$, $\dist(\Csig,\Csig') > \sigma$,
	where \smash{$\dist(\Csig,\Csig') := \inf_{x \in \Csig} \dist(x,\Csig')$}.

	\item 
	\label{asmp: low_noise_density}
	\emph{Low noise density:} There exists $\gamma,c_0 > 0$ such that for all $x
	\in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
	$$
	\inf_{x' \in \Csig} f(x') - f(x) \geq  c_0 \dist(x, \Csig)^{\gamma}.
	$$

	%%% AJG 5/20: Should I turn this into two conditions?
	\item
	\label{asmp: embedding}
	\emph{Lipschitz embedding:}
	There exists $g: \Reals^d \to \Reals^d$ which has the following properties: i) there exists a convex set $\mathcal{K} \subseteq \Rd$ with $\mathrm{diam}(\mathcal{K}) = \sup_{x,y \in \mathcal{K}}\norm{x - y} =: \diam < \infty$, such that $\Csig = g(\mathcal{K})$, ii) $\det(\nabla g (x)) = 1$ for all $x \in \Csig$, where $\nabla g(x)$ is the Jacobian of $g$ evaluated at $x$, and iii) for some $L \geq 1$, 
	\begin{equation*}
	\frac{1}{L}\norm{x - y} \leq \norm{g(x) - g(y)} \leq L \norm{x - y} ~ \text{for all $x,y \in \mathcal{K}$}.
	\end{equation*}
	Simply put, $\Csig$ is the image of a convex set with finite diameter, under a  measure preserving, biLipschitz transformation.
	\item
	\label{asmp: bounded_volume}
	\emph{Bounded volume:}
	Let the neighborhood graph radius $0 < r \leq \sigma/2d$ be such that
	\begin{equation*}
	2 \int_{\Csig} \Pbb(B(x,r)) f(x) dx \leq \int_{\Rd} \Pbb(B(x,r)) f(x) dx
	\end{equation*}
\end{enumerate}

To motivate these conditions, for an arbitrary graph $G = (V,E)$ and subset of vertices $S \subseteq V$, consider the normalized cut $\Phi(S;G)$ (defined as in \eqref{eqn: normalized_cut}), as well as the mixing time of a random walk over the induced subgraph $G[S]$ (defined in Section \ref{sec: analysis} by \eqref{eqn: mixing_time}, and denoted by $\tau_{\infty}(G[S])$). In \citet{zhu2013}, it is shown that for a constant $c > 0$, the \pprspace estimate $\Cest$ of $S$ satisfies
\begin{equation}
\label{eqn: graph_symmetric_set_difference_1}
\vol(\Cest \setminus S; G) + \vol(S \setminus \Cest; G) \leq c \bigl( \Phi(S,G) \cdot \tau_{\infty}(G[S])\bigr) \vol(S;G)
\end{equation}
where the left hand side resembles a (degree-weighted) form of the symmetric set difference of \eqref{eqn: misclassification_rate}.

As we will formally show in Section \ref{sec: analysis}, the conditions \ref{asmp: bounded_density}-\ref{asmp: bounded_volume} allow us to upper bound the normalized cut $\Phi(\Csig[\Xbf]; G_{n,r})$, and the mixing time $\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$. Intuitively, the low noise \ref{asmp: cluster_separation} and cluster separation \ref{asmp: low_noise_density} assumptions yield an upper bound on $\cut(\Csig[\Xbf]; G_{n,r})$, the lower bound on density in \ref{asmp: bounded_density} yields a lower bound on $\vol(\Csig[\Xbf]; G_{n,r})$, and along with \ref{asmp: bounded_volume}, which ensures $\vol(\Csig[\Xbf]; G_{n,r}) \leq \vol(\Csig[\Xbf]^c; G_{n,r})$, these imply an upper bound on the normalized cut. \ref{asmp: bounded_density} and \ref{asmp: embedding} preclude bottlenecks in the induced subgraph $G_{n,r}[\Csig[\Xbf]]$, and combined with the upper bound on diameter in \ref{asmp: embedding}, they yield an upper bound on the mixing time over this subgraph.

\paragraph{Condition number} We will return to the topic of conditions in Section \ref{sec: analysis}. Now, we define the condition number, $\kappa(\Cset)$, which reflects the difficulty of the local spectral clustering task. Motivated by \eqref{eqn: graph_symmetric_set_difference_1}, we will set $\kappa(\Cset)$ to be an upper bound on $\Phi(\Csig[\Xbf]; G_{n,r}) \cdot \tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$. The smaller $\kappa(\Cset)$ is, the more success \pprspace will have in recovering $\Cset$. Let $\theta := (r, \sigma, \lambda, \lambda_{\sigma}, \Lambda_{\sigma}, \gamma, \diam, L)$ contain those geometric parameters detailed in \ref{asmp: bounded_density} - \ref{asmp: bounded_volume}.

\begin{definition}[Well-conditioned density clusters]
	For $\lambda > 0$ and $\Cset \in \Cbb_f(\lambda)$, let $\Cset$ satisfy \ref{asmp: bounded_density} - \ref{asmp: bounded_volume} for some $\theta$. Then, for universal constants $c_1, c_2, c_3 > 0$ to be specified later, we set
	\begin{equation}
	\label{eqn: condition_number_1}
	\Phibf(\theta) 
	:= c_1 r \frac{d}{\sigma} \frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} - c_0 \frac{r^{\gamma}}{\gamma + 1})}{\lambda_{\sigma}},~ 
	\taubf(\theta) := c_2 \frac{\Lambda_{\sigma}^4 d^3 \rho^2 L^2}{\lambda_{\sigma}^4 r^2} \log^2\left(\frac{1}{r}\right) + c_3
	\end{equation}
	and letting $\kappa(\Cset) := \Phi_{u}(\theta) \cdot \tau_{u}(\theta)$, we call $\Cset$ a \textrm{$\kappa$-well-conditioned density cluster}.
\end{definition}

We note that $\Phibf(\theta)$ and $\taubf(\theta)$ are exactly the upper bounds on $\Phi(\Csig[\Xbf]; G_{n,r})$ and $\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$ that we derive in Section \ref{sec: analysis}.

\paragraph{Well-initialized algorithm}

As is typical in the local clustering literature, our algorithmic results will be stated with respect to specific ranges of each of the user-specified parameters.

In particular, for a well-conditioned density cluster $\Cset$ (with respect to some $\theta$), we require
\begin{align}
\label{eqn: initialization}
0 < r \leq \frac{\sigma}{2d}, & ~\alpha \in [1/10, 1/9] \cdot \frac{1}{\taubf(\theta)} \nonumber,  \\
v \in \Csig[\Xbf]^g, & ~\vol_0 \in [3/4,5/4] \cdot n(n-1) \int_{\Csig} \Pbb(B(x,r)) f(x) dx
\end{align}
where $\Csig[\Xbf]^g \subseteq \Csig[\Xbf]$ will be some large subset of $\Csig[\Xbf]$. In particular, letting $\vol_{n,r}(S) := \vol(S; G_{n,r})$ for $S \subseteq \Xbf$, we have $\vol_{n,r}(\Csig[\Xbf]^g) \geq \vol_{n,r}(\Csig[\Xbf])/2$.

@@

In practice it is clearly not feasible to set hyperparameters based on the underlying (unknown) density $f$. Typically, one tunes \pprspace over a range of hyperparameters and optimizes for some criterion such as minimum normalized cut; it is not obvious how this scheme would affect the performance of \pprspace in the density clustering context.

\paragraph{Density cluster estimation by \ppr}

The results of Section \ref{sec: analysis}, along with \eqref{eqn: graph_symmetric_set_difference_1}, give an upper bound on the volume of $\Cest \setminus \Csig[\Xbf]$ and $\Csig[\Xbf] \setminus \Cest$,
\begin{equation}
\label{eqn: graph_symmetric_set_difference}
\vol_{n,r}(\Cest \setminus \Csig[\Xbf]) + \vol_{n,r}(\Csig[\Xbf] \setminus \Cest) \leq c \kappa(\Cset) \vol_{n,r}(\Csig[\Xbf]).
\end{equation}
To translate \eqref{eqn: graph_symmetric_set_difference} into meaningful bounds on the symmetric set difference $\Delta(\Csig[\Xbf], \Cest)$, we wish to preclude vertices $x \in \Xbf$ from having arbitrarily small degree, and so we make some regularity assumptions on $\mathcal{X} := \mathrm{supp}(f)$. Let $\nu$ denote the Lebesgue measure on $\Rd$, and $\nu_d := \nu(B)$ be the measure of the unit ball $B = B(0,1)$.
\begin{enumerate}[label=(A\arabic*)]
	\setcounter{enumi}{5}
	\item 
	\label{asmp: valid_region}
	\emph{Regular support:} There exists some number $\lambda_{\min} > 0$ such that $\lambda_{\min} < f(x)$ for all $x \in \mathcal{X}$. Additionally, there exists some $c > 0$ such that for each $x \in \partial \mathcal{X}$, $\nu(B(x,r) \cap \mathcal{X}) \geq c \nu_d r^d$.
\end{enumerate}
Note that the latter condition in \ref{asmp: valid_region} will be satisfied if, for instance, the support $\mathcal{X}$ is a $\sigma$-expanded set.

\begin{theorem}
	\label{thm: misclassification_rate}
	Fix $\lambda > 0$, let $\Cset \in \Cbb_f(\lambda)$ be a $\kappa$-well-conditioned density cluster (with respect to some $\theta$), and additionally assume $f$ satisfies \ref{asmp: valid_region}. Then, there exists a universal constant $c_4 > 0$ such that with probability tending to $1$ as $n \to \infty$, 
	\begin{equation}
	\label{eqn: misclassification_rate_ub}
	\Delta(\Csig[\Xbf], \Cest) \leq c_4 \kappa(\Cset) \frac{\Lambda_{\sigma}}{\lambda_{\min}}.
	\end{equation}
\end{theorem}

The proof of Theorem \ref{thm: misclassification_rate}, along with all other proofs in this paper, can be found in the supplementary material. We observe that the symmetric set difference $\Delta(\Csig[\Xbf], \Cest)$  is proportional to the difficulty of the clustering problem, as measured by the condition number $\kappa(\Cset)$.

Neither \eqref{eqn: graph_symmetric_set_difference} nor Theorem \ref{thm: misclassification_rate} imply consistent density cluster estimation in the sense of \eqref{eqn: consistent_density_cluster_recovery}. This notion of consistency requires a uniform bound over $\pbf$: namely, for all $\Cset' \in \Cbb_f(\lambda), \Cset' \neq \Cset$, and each $u \in \Cset, w \in \Cset'$,
\begin{equation}
\label{eqn: ppr_gap}
\frac{p_{w}}{\Dbf_{ww}} \leq \frac{1}{40\vol_0} < \frac{1}{11\vol_0} \leq \frac{p_u}{\Dbf_{uu}},
\end{equation}
so that any sweep cut $S_{\beta}$ for $\beta \vol_0 \in [1/40,1/11]$ (i.e. any sweep cut considered by Algorithm \ref{alg: ppr}) will fulfill both conditions laid out in \eqref{eqn: consistent_density_cluster_recovery}. In Theorem \ref{thm: consistent_recovery_of_density_clusters}, we show that a sufficiently small upper bound on $\kappa(\Cset)$ ensures such a gap exists with probability one as $n \to \infty$, and therefore guarantees $\Cest$ will be a consistent estimator. As was the case before, we wish to preclude arbitrarily low degree vertices, this time for points $x \in \Cset'[\Xbf]$.
\begin{enumerate}[label=(A\arabic*)]
	\setcounter{enumi}{6}
	\item 
	\label{asmp: C'_bounded_density}
	\emph{Bounded density:} Letting $\sigma,\lambda_{\sigma}$ be as in \ref{asmp: bounded_density}, for each $\Cset' \in \Cbb_f(\lambda)$ and for all $x \in \Csig'$, $\lambda_{\sigma} \leq f(x)$.
\end{enumerate}

\begin{theorem}
	\label{thm: consistent_recovery_of_density_clusters}
	Fix $\lambda > 0$, let $\Cset \in \Cbb_f(\lambda)$ be a $\kappa$-well-conditioned cluster (with respect to some $\theta$), and additionally assume $f$ satisfies \ref{asmp: C'_bounded_density}. If Algorithm \ref{alg: ppr} is well-initialized, there exists a universal constant $c_5 > 0$ such that if
	\begin{equation}
	\label{eqn: kappa_ub}
	\kappa(\Cset) \leq c_5 \frac{\lambda_{\sigma}^2r^d \nu_d}{\Lambda_{\sigma}\Pbb(\Csig)},
	\end{equation}
	then the output set $\Cest \subseteq \Xbf$ is a consistent estimator for $\Cset$, in the sense of Definition \ref{def: consistent_density_cluster_estimation}.
\end{theorem}

\begin{remark}
	We note that the restriction on $\kappa(\Cset)$ imposed by \eqref{eqn: kappa_ub} results in a symmetric set difference $\Delta(\Csig[\Xbf], \Cest)$ on the order of $r^d$. In plain terms, we are able to recover a density cluster $\Cset$ in the sense of \eqref{eqn: consistent_density_cluster_recovery} only when we can guarantee a very small fraction of points will be misclassified. This strong condition is the price we pay in order to obtain the uniform bound of \eqref{eqn: ppr_gap}.
\end{remark}

\begin{remark}
	While taking the radius of the neighborhood graph $r \to 0$ as $n \to \infty$---and thereby ensuring $G_{n,r}$ is sparse---is computationally attractive, the presence of a factor of $\frac{\log^2(1/r)}{r}$ in $\kappa(\Cset)$ unfortunately prevents us from making claims about the behavior of \pprspace in this regime. Although the restriction to a kernel function fixed in $n$ is standard for theoretical analysis of spectral clustering \cite{schiebinger2015,vonluxburg2008}, it is an interesting question whether \pprspace exhibits some degeneracy over $r$-neighborhood graphs as $r \to 0$, or if this is merely looseness in our upper bounds.
\end{remark}

>>>>

When applied to geometric graphs built from a large number of samples,
global spectral clustering methods can be computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to increased interest in
local spectral clustering algorithms, which leverage locally-biased spectra
computed using random walks around some user-specified seed node.  A popular 
local clustering algorithm is Personalized PageRank (PPR), first introduced by 
\citet{haveliwala2003}, then further developed by
\citep{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013},
among others.

@@

\paragraph{Estimation of density clusters.} Let \smash{$\Cbb_f(\lambda)$} denote 
the connected components of the density upper level set $\{x \in \Rd: f(x) >
\lambda\}$.  For a given density cluster \smash{$\Cset \in \Cbb_f(\lambda)$}, we
call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical density cluster}. The
size of the symmetric set difference between estimated and empirical clusters is 
a commonly used metric to quantify cluster estimation error
\citep{korostelev1993,polonik1995,rigollet2009}.

@@

However, the symmetric set difference does not measure whether \smash{$\Cest$} 
can distinguish any two distinct clusters \smash{$\Cset,\Cset' \in
  \Cbb_f(\lambda)$}. We therefore also study a second notion of cluster
estimation, first introduced by \citet{hartigan1981}, and defined
asymptotically.

@@

More broadly, density clustering and level set estimation is a well-studied
problem. \citet{polonik1995, rigollet2009} study density clustering under the 
symmetric set difference metric, \citet{tsybakov1997, singh2009} describe
minimax optimal level set estimators under Hausdorff loss and
\citet{hartigan1981, chaudhuri2010} consider consistent estimation of the
cluster tree, to note but a few works. Our goal is not to improve on these
results, nor to offer a better algorithm for level set estimation; indeed, seen as
a density clustering algorithm, PPR has none of the optimality guarantees 
found in the aforementioned works. Instead, our motivation is start with a
widely-used local spectral method, PPR, and to better understand and
characterize the distinctions between those density clusters which are
well-conditioned for PPR, and those which are not.

@@

We formalize some geometric conditions, and use these to define a condition
number $\kappa(\Cset)$, which measures the difficulty PPR will have in  
estimating $\Cset$. (Our theoretical guarantees for PPR will be framed in terms 
of $\kappa(\Cset)$.)

\paragraph{Geometric conditions on density clusters.} At a high level, for PPR
to be successful, the underlying density cluster must be geometrically
well-conditioned.  At a minimum, we want to avoid sets that contain arbitrarily
thin bridges or spikes.  Hence, as in \cite{chaudhuri2010}, we consider a
thickened version of \smash{$\Cset \in \Cbb_f(\lambda)$} defined as 
$\Csig := \set{x \in \Reals^d: \dist(x,\Cset) \leq \sigma}$, which 
we call the \emph{$\sigma$-expansion} of $\Cset$. Here 
\smash{$\dist(x,\Cset) := \inf_{y \in \Cset} \norm{y - x}$}.  We now list our
conditions on $\Csig$.

\begin{enumerate}[label=(A\arabic*)]
\item
  \label{asmp: bounded_density}
  \emph{Bounded density within cluster:} There exist constants
  $0<\lambda_{\sigma}< \Lambda_{\sigma}<\infty$ such that 
  $\lambda_{\sigma} \leq \inf_{x \in \Csig} f(x) \leq \sup_{x \in \Csig} f(x)
  \leq \Lambda_{\sigma}$.

\item
  \label{asmp: cluster_separation}
  \emph{Cluster separation:}
  For all clusters $\Cset' \in \Cbb_f(\lambda)$ with $\Cset' \not= \Cset$, 
  $\dist(\Csig,\Csig') > \sigma$, where $\dist(\Csig,\Csig') := \inf_{x
    \in \Csig} \dist(x,\Csig')$.

\item 
  \label{asmp: low_noise_density}
  \emph{Low noise density:} There exist $\gamma,c_0 > 0$ such that for 
  $x \in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
  $\inf_{x' \in \Csig} f(x') - f(x) \geq  c_0 \dist(x, \Csig)^{\gamma}$.

  %%% AJG 5/20: Should I turn this into two conditions?
\item
  \label{asmp: embedding}
  \emph{Lipschitz embedding:}
  There exists $g: \Reals^d \to \Reals^d$ with the following properties: i)
  we have $\Csig = g(\mathcal{K})$, for a convex set $\mathcal{K} \subseteq \Rd$
  with $\mathrm{diam}(\mathcal{K}) = \sup_{x,y \in \mathcal{K}}\norm{x - y} =:
  \diam < \infty$; ii) $\det(\nabla g (x)) = 1$ for all $x \in \Csig$, where
  $\nabla g(x)$ is the Jacobian of $g$ evaluated at $x$; and iii) for some $L
  \geq 1$,   
  \begin{equation*}
    \frac{1}{L}\norm{x - y} \leq \norm{g(x) - g(y)} \leq L \norm{x - y} ~
    \text{for all $x,y \in \mathcal{K}$}. 
  \end{equation*}
  Succintly, $\Csig$ is the image of a convex set with finite diameter 
  under a  measure preserving, bi-Lipschitz transformation.

\item
  \label{asmp: bounded_volume}
  \emph{Bounded volume:}
  Let the neighborhood graph radius $0 < r \leq \sigma/2d$ be such that
  \begin{equation*}
    2 \int_{\Csig} \Pbb(B(x,r)) f(x) dx \leq \int_{\Rd} \Pbb(B(x,r)) f(x) dx,
  \end{equation*}
  where $B(x,r)$ is the closed ball of radius $r$ at $x$.
\end{enumerate}

To motivate these conditions, \citet{zhu2013} show for arbitrary graph $G =
(V,E)$ and subset of vertices $S \subseteq V$, the PPR estimate \smash{$\Cest$}
of subset $S$ satisfies, for a constant $c>0$,
\begin{equation}
\label{eqn: graph_symmetric_set_difference_1}
\vol(\Cest \setminus S; G) + \vol(S \setminus \Cest; G) \leq c \bigl( \Phi(S,G)
\cdot \tau_{\infty}(G[S])\bigr) \vol(S;G),
\end{equation}
where $\Phi(S;G)$ is the normalized cut of $S$ (as defined in \eqref{eqn:
  normalized_cut}), and \smash{$\tau_{\infty}(G[S])$} is called the \emph{mixing
  time} of a random walk over the induced subgraph $G[S]$ (to be defined
precisely later, in \eqref{eqn: mixing_time}).  The left-hand side in
\eqref{eqn:  graph_symmetric_set_difference_1} resembles a (degree-weighted) 
form of the symmetric set difference metric in \eqref{eqn:
  misclassification_rate}.  As we will show in Section \ref{sec:
  analysis}, the conditions \ref{asmp: bounded_density}-\ref{asmp:
  bounded_volume} allow us to upper bound the normalized cut 
\smash{$\Phi(\Csig[\Xbf]; G_{n,r})$}, and the mixing time 
\smash{$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$}.  (Informally, 
\ref{asmp: cluster_separation}, \ref{asmp: low_noise_density} yield an upper
bound on \smash{$\cut(\Csig[\Xbf]; G_{n,r})$}, and \ref{asmp: bounded_density}
yields a lower bound on \smash{$\vol(\Csig[\Xbf]; G_{n,r})$}; together with 
\ref{asmp: bounded_volume}, this gives an upper bound on the normalized cut.  On 
the other hand, \ref{asmp: bounded_density}, \ref{asmp: embedding} preclude
bottlenecks in the induced subgraph \smash{$G_{n,r}[\Csig[\Xbf]]$}, and combined
with \ref{asmp: embedding}, this leads to an upper bound on the mixing time over 
this subgraph.)

\paragraph{Condition number.} Motivated by \eqref{eqn:
  graph_symmetric_set_difference_1}, we will define $\kappa(\Cset)$ to be an
upper bound on the product \smash{$\Phi(\Csig[\Xbf]; G_{n,r}) \cdot
  \tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$}. The smaller $\kappa(\Cset)$ is, the
more success PPR will have in recovering $\Cset$. Let $\theta := (r, \sigma,
\lambda, \lambda_{\sigma}, \Lambda_{\sigma}, \gamma, \diam, L)$ contain the
geometric parameters from \ref{asmp:  bounded_density}--\ref{asmp:
  bounded_volume}.

\begin{definition}[Well-conditioned density clusters]
  For $\lambda > 0$ and \smash{$\Cset \in \Cbb_f(\lambda)$}, let $\Cset$ satisfy 
  \ref{asmp: bounded_density}--\ref{asmp: bounded_volume} for some
  $\theta$. Then, for universal constants $c_1, c_2, c_3 > 0$ to be specified
  later, we set 
  \begin{equation}
    \label{eqn: condition_number_1}
    \Phibf(\theta) 
    := c_1 r \frac{d}{\sigma} \frac{\lambda}{\lambda_{\sigma}}
    \frac{(\lambda_{\sigma} - c_0 \frac{r^{\gamma}}{\gamma +
        1})}{\lambda_{\sigma}},~  
    \taubf(\theta) := c_2 \frac{\Lambda_{\sigma}^4 d^3 \rho^2
      L^2}{\lambda_{\sigma}^4 r^2} \log^2\left(\frac{1}{r}\right) + c_3,
  \end{equation}
  and letting \smash{$\kappa(\Cset) := \Phi_{u}(\theta) \cdot
    \tau_{u}(\theta)$}, we call $\Cset$ a 
  \emph{$\kappa$-well-conditioned density cluster}.  
\end{definition}

We note that $\Phibf(\theta)$ and $\taubf(\theta)$ are exactly the upper bounds 
on \smash{$\Phi(\Csig[\Xbf]; G_{n,r})$} and
\smash{$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$} that we derive in our analysis
later, in Section \ref{sec: analysis}.

\paragraph{Well-initialized algorithm.} As is typical in the local clustering
literature, our algorithmic results will be stated with respect to specific
ranges of each of the user-specified parameters. In particular, for a
well-conditioned density cluster $\Cset$ (with respect to some $\theta$), we
require 
\begin{equation}
\begin{gathered}
\label{eqn: initialization}
0 < r \leq \frac{\sigma}{2d}, \quad 
\alpha \in [1/10, 1/9] \cdot \frac{1}{\taubf(\theta)}, \\
v \in \Csig[\Xbf]^g, \quad
\vol_0 \in [3/4,5/4] \cdot n(n-1) \int_{\Csig} \Pbb(B(x,r)) f(x) dx, 
\end{gathered}
\end{equation}
where \smash{$\Csig[\Xbf]^g \subseteq \Csig[\Xbf]$} will be some large
(``good'') subset of $\Csig[\Xbf]$. In particular, abbreviating $\vol_{n,r}(S)
:= \vol(S; G_{n,r})$ for $S \subseteq \Xbf$, we will have
\smash{$\vol_{n,r}(\Csig[\Xbf]^g) \geq \vol_{n,r}(\Csig[\Xbf])/2$}.

@@

In practice it is clearly not feasible to set hyperparameters based on the
underlying (unknown) density $f$. Typically, one tunes PPR over a range of
hyperparameters and optimizes for some criterion such as minimum normalized cut;
it is not obvious how this scheme would affect the performance of PPR in the
density clustering context.

\paragraph{Main theorems.} The results of Section \ref{sec: analysis}, combined 
with \eqref{eqn: graph_symmetric_set_difference_1}, give an upper bound on the
volume of \smash{$\Cest \setminus \Csig[\Xbf]$} and \smash{$\Csig[\Xbf]
  \setminus \Cest$},  
\begin{equation}
\label{eqn: graph_symmetric_set_difference}
\vol_{n,r}(\Cest \setminus \Csig[\Xbf]) + \vol_{n,r}(\Csig[\Xbf] \setminus
\Cest) \leq c \kappa(\Cset) \vol_{n,r}(\Csig[\Xbf]). 
\end{equation}
To translate \eqref{eqn: graph_symmetric_set_difference} into meaningful bounds
on the symmetric set difference metric \smash{$\Delta(\Csig[\Xbf], \Cest)$}, we
want to preclude vertices $x \in \Xbf$ from having arbitrarily small degree, and
so we make some regularity assumptions on $\mathcal{X} := \mathrm{supp}(f)$. Let
$\nu$ denote the Lebesgue measure on $\Rd$, and $\nu_d := \nu(B)$ be the measure
of the unit ball $B = B(0,1)$. 
\begin{enumerate}[label=(A\arabic*)]
  \setcounter{enumi}{5}
\item 
  \label{asmp: valid_region}
  \emph{Regular support:} There exists some constant $\lambda_{\min} > 0$ such 
  that $\lambda_{\min} < f(x)$ for all $x \in \mathcal{X}$. Additionally, there
  exists some $c > 0$ such that for each $x \in \partial \mathcal{X}$,
  $\nu(B(x,r) \cap \mathcal{X}) \geq c \nu_d r^d$. 
\end{enumerate}
Note that the latter condition in \ref{asmp: valid_region} will hold if the
boundary $\partial \mathcal{X}$ is sufficiently regular.  Now we present our
main bound on the symmetric set difference metric.

\begin{theorem}
  \label{thm: misclassification_rate}
  Fix $\lambda > 0$, let \smash{$\Cset \in \Cbb_f(\lambda)$} be a
  $\kappa$-well-conditioned density cluster (with respect to some $\theta$), and
  additionally assume $f$ satisfies \ref{asmp: valid_region}. If Algorithm
  \ref{alg: ppr} is well-initialized, there exists a universal constant $c_4 >
  0$ such that with probability tending to $1$ as $n \to \infty$,  
  \begin{equation}
    \label{eqn: misclassification_rate_ub}
    \Delta(\Csig[\Xbf], \Cest) \leq c_4 \kappa(\Cset)
    \frac{\Lambda_{\sigma}}{\lambda_{\min}}. 
  \end{equation}
\end{theorem}

The proof of Theorem \ref{thm: misclassification_rate}, along with all other
proofs in this paper, is deferred to the supplementary material. Note that this 
result says the symmetric set difference metric \smash{$\Delta(\Csig[\Xbf],
  \Cest)$} is proportional to the difficulty of the clustering problem, as
measured by the condition number $\kappa(\Cset)$.

Neither \eqref{eqn: graph_symmetric_set_difference} nor \eqref{eqn:
  misclassification_rate_ub} imply consistent density cluster estimation in the 
sense of \eqref{eqn: consistent_density_cluster_recovery}. This notion of
consistency requires a uniform bound over $\pbf$: for all \smash{$\Cset'
  \in \Cbb_f(\lambda), \Cset' \neq \Cset$}, and each $u \in \Cset, w \in
\Cset'$,  
\begin{equation}
\label{eqn: ppr_gap}
\frac{p_{w}}{\Dbf_{ww}} \leq \frac{1}{40\vol_0} < \frac{1}{11\vol_0} \leq
\frac{p_u}{\Dbf_{uu}}, 
\end{equation}
so that any sweep cut $S_{\beta}$ for $\beta \vol_0 \in [1/40,1/11]$ (i.e., any
sweep cut considered by Algorithm \ref{alg: ppr}) will fulfill both conditions
laid out in \eqref{eqn: consistent_density_cluster_recovery}. In Theorem
\ref{thm: consistent_recovery_of_density_clusters}, we show that a sufficiently
small upper bound on $\kappa(\Cset)$ ensures such a gap exists with probability
1 as $n \to \infty$, and hence guarantees \smash{$\Cest$} will be a consistent  
estimator. As was the case before, we wish to preclude arbitrarily low degree
vertices, this time for points $x \in \Cset'[\Xbf]$. 
\begin{enumerate}[label=(A\arabic*)]
  \setcounter{enumi}{6}
\item 
  \label{asmp: C'_bounded_density}
  \emph{Bounded density in other clusters:} Letting $\sigma,\lambda_{\sigma}$ be 
  as in \ref{asmp: bounded_density}, for each $\Cset' \in \Cbb_f(\lambda)$ and
  for all $x \in \Csig'$, $\lambda_{\sigma} \leq f(x)$. 
\end{enumerate}

Next we give our main result on consistent cluster recovery by PPR.

\begin{theorem}
  \label{thm: consistent_recovery_of_density_clusters}
  Fix $\lambda > 0$, let $\Cset \in \Cbb_f(\lambda)$ be a
  $\kappa$-well-conditioned density cluster (with respect to some $\theta$), and
  additionally assume $f$ satisfies \ref{asmp: C'_bounded_density}. If Algorithm
  \ref{alg: ppr} is well-initialized, there exists a universal constant $c_5 >
  0$ such that if 
  \begin{equation}
    \label{eqn: kappa_ub}
    \kappa(\Cset) \leq c_5 \frac{\lambda_{\sigma}^2r^d
      \nu_d}{\Lambda_{\sigma}\Pbb(\Csig)}, 
  \end{equation}
  then the output set $\Cest \subseteq \Xbf$ is a consistent estimator for
  $\Cset$, in the sense of Definition \ref{def: consistent_density_cluster_estimation}. 
\end{theorem}

\begin{remark}
  We note that the restriction on $\kappa(\Cset)$ imposed by \eqref{eqn:
    kappa_ub} results in a symmetric set difference \smash{$\Delta(\Csig[\Xbf],
    \Cest)$} on the order of $r^d$. In plain terms, we are able to recover a
  density cluster $\Cset$ in the sense of \eqref{eqn:
    consistent_density_cluster_recovery} only when we can guarantee a very small
  fraction of points will be misclassified. This strong condition is the price
  we pay in order to obtain the uniform bound of \eqref{eqn: ppr_gap}. 
\end{remark}
